The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}

### 2014-02-11 article/article.tex
review

\usepackage{color}
\newcommand\todo[1]{\textcolor{red}{(#1)}}

@@

\title{The Power of Two Random Choices\\ 
\large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Träff, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}}
\author{Martin Kalany, 0825673}

\maketitle

\begin{abstract}

@@

%\tableofcontents

\section{Introduction}
\label{sec:intro}
Suppose that $m$ balls are thrown into $n$ bins, according to some strategy. We are interested in the \emph{number of balls in the fullest bin}, also called the \emph{maximum load}. Equivalently, one can look at the \emph{addative gap} of the allocation, that is the difference of the number of balls in the fullest bin and the average amount of balls in the bins. These \emph{balls-into-bins games} or \emph{allocation problems} have been studied extensively in the probability literature (see e.g., \cite{JK77}). In the context of Computer Science, the balls-into-bins model is useful to study e.g., strategies for distributing \emph{tasks} (balls) to \emph{processors} (bins) as evenly as possible without having to query all processors for the number of tasks currently assigned to them.

A simple strategy for the balls-into-bins problem is to choose one bin uniformly and independently at random for each ball, which we will call the \emph{single-choice balls-into-bins} strategy. Using this strategy, the maximum load of the fullest bin will be $\frac{m}{n} + \Theta\left(\sqrt{\frac{m \cdot \log n}{n}}\right)$ with high probability\footnote{We say that an event $\mathcal E$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E \right] = 1 - o(1)$, except when stated otherwise.}~\cite{RS98}.

Consider a variation of the above \emph{single-choice balls-into-bins} strategy: For each ball, chose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}\footnote{If there are multiple bins with the same minimal load, chose one among them arbitrarily.}. This simple \emph{multiple-choice balls-into-bins} scheme, which we will call the \emph{Greedy} algorithm, achieves a maximum load of $\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d}\right)$ w.h.p.~\cite{ABKU99}~\cite{BCSV06}. This apparently minor change introduces a small amount of choice which, compared to the single-choice scheme, leads to a significant reduction in the maximum load. 
\begin{comment}
Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. 
\end{comment}
Extensive research following the influental paper by Azar et. al \cite{ABKU99}, who were the first to explicitly prove this phenomenon, suggests that this \emph{two-choice paradigm} is not just an artificat of the simple balls-into-bins model, but a general principle applicable to a variety of problems.

Condsider a slightly more involved strategy, which we will call the \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$ \todo{simplify to n/d?}. We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with same minimal load, we put the ball into the bin with smallest index $i$. Somewhat surprisingly, V\"ocking \cite{VOC03} showed that such a strategy employing \emph{asymetric tie-breaking} in combination with \emph{partitioning}\todo{check} reduces the bound for the gap to $\Theta \left( \frac{\log\log n}{d  \cdot \log \Phi_d}\right)$ \todo{$\Phi_d$ is undefined}.

The paper is organized as follows: We will first introduce different variations of the \emph{balls-into-bins problem} in Section \ref{sec:classification} to provide some overview of different directions in research as well as possibilities of abstraction from applications, some of which are listed in Section \ref{sec:applications}. Sections \ref{sec:single-choice} to \ref{sec:multiple-choice} will deal with the most basic variation of the balls-into-bins problem, namely the sequential, finite and static variant (see Section \ref{sec:classification} for details). For comparison and as a starting point, Section \ref{sec:single-choice} provides the results for the single-choice balls-into-bins problem. \todo{TODO: multiple-choice, analysis, mention that witness tree method very helpful}  See also \cite{MRS01} an overview of proof techniques for the given problem.

\begin{comment}
In Section \ref{sec:strategies}, we will provide asymptotic bounds for the addative gap of this strategy, which we will then use to devise more complex strategies that significantly reduce these bounds. 
\end{comment}

@@

\section{Classification}
\label{sec:classification}
Several different variations of this problem exist, which may be classified by the following properties:
\begin{compactitem}
\item In a \emph{sequential process}, the balls arrive one after another. A  ball is dealt with only after the previous ball has been placed. Contrary, in a \emph{parallel process}, balls arrive in sets and the balls in one set are dealt with simultaneously \cite{ABS98}. 
\item In a \emph{static problem}, the balls are placed into the bins and no balls are ever removed. In a \emph{dynamic problem}, balls can also be removed from the system. In a simple \emph{random deletions} model, for each newly inserted ball a ball choosen uniformly and independently at random from among the balls currently in the bins is removed from the system \cite{ABKU99} \cite{MRS01}. More elaborate shemes exists: An \emph{oblivious adversery} specifies the sequence of insertions and deletions of balls in advance, without any knowledge of the placement of the balls \cite{CFM+98}. Another approach was studied by Adler et. al \cite{ABS98} in the context of the parallel balls-into-bins problem: In each bin, the balls are stored in FIFO-order and at every time step the first ball in each of the bins is deleted. 
\item A \emph{finite process} is analyzed for a given time interval that is known beforehand, whereas for an \emph{infinite process}, a placement and a deletion process is defined over an infinite time span \cite{ABS98}.
\end{compactitem}

We will distinguish two cases with respect to the number of balls $m$. In the \emph{lightly loaded case}, $m$ is in the order of $n$, whereas in the \emph{heavily loaded case}, $m$ is aribtrarily large. For clarity, we will assume that $m = n$ in the slightly loaded case and that $m \ll n$\todo{TODO} \footnote{Let $f$, $g$ be arbitrary functions. Then, $f \ll g$ means $f = o(g)$ and $ f \gg g$ means $f = \omega(g)$} in the heavily loaded case.

@@

To maximize the throughput of the servers, a uniform distribution of tasks to servers is desireable. This can easily be achieved by a central dispatcher. However, such a central dispatcher poses a bottleneck within a distributed system and querying $m$ servers for their current load involves significant overhead, e.g., sending a message to each server and waiting for the reply and more efficient solutions are neccessary. As with hashing, the balls-into-bins model can be applied easily: we need to assign tasks (the balls) to servers (the bins). Assuming tasks are assigned to servers uniformly and independently at random, the two-choice paradigm is directly applicable: If each client samples two random servers for their load and sends issues its request to the least loaded server, the overhead is comperatively small and the difference in load on the $n$ servers is bounded by $\Theta\left(\log \log n\right)$ w.h.p. \cite{KLM92} \cite{RS98} \cite{MRS01}.

\subsection{Emulation of shared memory machines on distributed memory machines}
\label{sec:DMM}
The \emph{Parallel Random Access Machine} model\footnote{as defined in the lecture or in~\cite{P03}}, or \emph{PRAM} for short, is a high level abstraction of a shared memory machine, in which memory access by any processor to any memory location is assumed to require constant time. It is especially useful for the study of parallel algorithms. However, a PRAM is much harder to realize in hardware than a \emph{Distributed Memory Machine} (DMM). Thus, ways to emulate a PRAM on a DMM have been studied extensivly. Such an emulation distributes the processors and memory cells of the PRAM to the processors and memory modules (typically one per processor) of the DMM with the goal to reduce the slowdown of the emulation, which is the time the DMM requires to simulate one step of the PRAM. Each memory modul can handle only one memory access at a time step. More accesses to the same module will result in memory contention. The balls-into-bins idea is useful for managing communication between the processors and the memory modules such that memory contention is minimal. For details, see Karp et. al \cite{KLM92} and the papers referenced there.

\subsection{Low congestion circuit routing}
\label{sec:circuitrouting}
In multi-stage interconnection networks, virtual circuit-switching is used to route communication. For simplicity, we consider the \emph{permutation routing problem}, where one request originates at each of the $n$ input nodes and one request is destined for each of the $n$ output nodes and assume a \emph{butterfly network} as the underlying interconnection network (see \cite{CLR09} for a detailed definition). Then, congestions is the maximum number of paths that are routed through a network link. The goal of a \emph{circuit routing algorithm} is to to allocate paths for all $n$ communication requests such that congestion is minimal.

@@

\section{Single-choice balls-into-bins}
\label{sec:single-choice}
We will now consider the most basic variant of the balls-into-bins problem: Namely the static, finite and sequential variant. Thus, the balls are assumed to arrive sequentially, no strategy for deleting balls in the bins has to be considered and we are interested in a bound for the maximum load or gap only after all $m$ balls have arrived.

In the \emph{single-choice balls-into-bins} strategy, each ball is placed into a bin chosen uniformly and independently at random. In other words, a ball gets placed into a certain bin with probability $\frac{1}{n}$.

The fullest bin contains
\begin{align}
\Theta \left( \frac{\log n}{\log \log n} \right)
\end{align} 
balls in the lightly loaded case w.h.p. Since the average load of the bins is 1, this is also a bound for the gap. 
%More accurately, the number of balls in the bin with the most balls is $\Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$ \cite{G91}. %\todo{really include Gonnet?}
In the \emph{heavily loaded case} the maximum load is 
\begin{align}
\frac{m}{n} + \Theta\left(\sqrt{\frac{m \cdot \log n}{n}}\right)
\end{align} 
w.h.p.~\cite{RS98}.

@@

Thus, let $B$ be the set of bins $\{1,2...n\}$, $\Omega = B^{d}$ be the sample space and $\mathcal{F} = 2^{\Omega}$ be the set of measurable events. We are only interested in events that represent a set of bins of size $d$. Thus, let $\mathcal{F}_d = \{\mathcal{E} \in \mathcal{F}: \left\vert \mathcal{E} \right\vert = d\}$ and set $\Pr\left(\mathcal{E}  \right) = 0$  $\forall \mathcal{E} \in \mathcal{F} \setminus \mathcal{F}_d $. The three classes of algorithms are distinguished by the probability function $\Pr\left(\mathcal{E}\right)$ for events $\mathcal{E} \in \mathcal{F}_d$:
\begin{compactitem}
\item  \emph{Uniform and independent.} Choose each of the $d$ bins for a ball uniformly and independently at random from the set of bins $B$. This implies 
\begin{align}
\forall \mathcal{E} \in \mathcal{F}_d: \Pr\left(\mathcal{E}\right) = \frac{1}{n ^{d}}
\end{align}
\item Class 2: \emph{Non-uniform and independent.} Choose the $i$th bin from $B$ independently at random according to some probability function
\begin{align}
\Pr: B \rightarrow \left[0,1\right]
\end{align}

@@

\todo{specify deletions; fix intro text since we now do use deletions}

\todo{use theorem from V\"ocking?}

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, at any time $t$ the maximum load is at most 
\begin{align}
\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d} \right)
\end{align}
w.h.p.~at any time $t$ \cite{ABKU99} \cite{BCSV06}.

\end{theorem}

@@

\subsubsection{History}
\label{sec:historyOfGreedy}
The history of proving this remarkable property of the two-choice paradigm is in itself quite interesting. It was first proven by Azar et al. \cite{ABKU99} for the case $m = n$. In the same paper, they claimed that the bound holds for arbitrary $m$. Unfortunately, their analysis breaks down for the heavily loaded case, where $m = \omega\left(n\right)$. Berenbrink at. all \cite{BFZR08} were the first to proof that indeed the same bound holds for the heavily loaded case\footnote{However, they use a slightly different definition for w.h.p.: An event $\mathcal E$ occurs with high probability if $\Pr\left[\mathcal E \right]  \geq 1- n^{-\alpha}$, for an arbitrarily chosen constant $\alpha \geq 1$.}. A simplified proof may be found in \cite{TW13}. Curiously, the case $d = 2$ was implicitly proven by Karp et al. in the context of simulating PRAMs on distributed memory machines\cite{KLM92}. As we will see in the next section, V\"ocking was able to slightly improve the bound as well as to show that the \emph{Always-go-left} strategy that he introduces is optimal up to additive constants \cite{VOC03}.

\subsection{The Always-go-left scheme}
\label{sec:AlwaysGoLeft}
The \emph{Greedy} strategy presented in Section \ref{sec:multiple-choice} assumes that $d$ bins are chosen uniformly and independently at random. V\"ocking \cite{VOC03} was the first to study the impact of non-uniform and dependent selections of the $d$ bins. Surprisingly, a multiple-choice balls-into-bins strategy with a \emph{non-uniform} and independent selection of the $d$ bins, such as the \emph{Always-go-left} algorithm introduced in \cite{VOC03}, achieves better\cite{BCSV06} load balancing than the \emph{Greedy} scheme (Section \ref{sec:greedy}). Furthermore, V\"ocking proved that the \emph{Always-go-left} algorithm is optimal up to additive constants.

@@

\begin{theorem}
\label{theorem:agln}
If $n$ balls are placed into $n$ bins using the Always-go-left scheme, the number of balls in the fullest bin will be 
\begin{align}
\frac{\ln\ln n}{d \cdot \ln \Phi_d} \pm \Theta(1)
\end{align}
w.h.p.
\end{theorem}

@@

When the maximum load exceeds some given threshold value we say that a \emph{bad event} occurs and construct an \emph{activated witness tree}. In other words, the activation of a witness tree is implied by a bad event. The probability for the existance of such an activated witness tree is an upper-bound for the probability that a bad event occurs.

Throughout this section, we donate the $d$ bins that are selected for a ball as the \emph{locations} of a ball. That is, before inserting a ball $d$ locations are generated and the ball is then inserted into one of those.

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections \ref{sec:analysis:greedy} and \ref{sec:analysis:alg}, respectively. However, we use a simplifying assumption \ref{assumption:independence} in those two sections to provide a more clear presentation of the proof technique. Section \ref{sec:analysis:nondistinctBalls} will extend these proofs to remove the simplifying assumption we make initially. Section \ref{sec:analysis:moreBalls} will extend the proof to cover the heavily loaded case where $m > n$ as well.

\subsection{Preliminaries}
\label{sec:preliminaries}
\todo{where to put this? required for  bound of always-go-left in intro}
We define $d$-ary Fibonacci numbers as follows. For $k \leq0$, $F_d(k) = 0$, $F_d(1) = 1$ and for $k \geq 1$,

\subsection{Greedy}
\label{sec:analysis:greedy}
\subsubsection{Symmetric witness trees}
\label{sec:analysis:definitionSymWT}
A \emph{symmetric witness tree} of order $L$ is a complete $d$-ary tree with $d^{L}$ leaves. Each node $v$ represents a ball $b_v = \mathrm{ball}(v)$ . The same ball may be represented by several nodes. The nodes and edges of the witness tree represent events that may occur or not. For simplicity, we assume that all events represented by a witness tree are stochastically independent.

We define edge and leaf events:
\begin{compactitem}
\item \emph{Edge event:} Consider some edge $e = (u,v)$ of the witness tree, s.t.~$v$ is the $i$th child of $u$. Then $e$ represents the event that the $i$'th location of the ball $b_u = \mathrm{ball}(u)$ points to the same bin as one of the locations of the ball $b_v = \mathrm{ball}(v)$. 
\todo{illustrate}
\item \emph{Leaf event:} A leaf node $v$ of the witness tree represents the event that each of the $d$ locations of the ball $b_v = \mathrm{ball}(v)$ points to a bin that contains at least three additional balls (i.e., balls that are not represented by a node of the tree) at the time of the insertion of the ball $b_v$.
\todo{illustrate}
\end{compactitem}

The following constraints have to be fullfilled by an assignment. The ball $b_r$ that is represented by the root node $r$ has to exist at timestep\footnote{As defined in Theorem \ref{theorem:greedy}} $t$ and each ball $b_v$ represented by a node $v \neq r$ and parent node $u$ has to exist when the ball $b_u = \mathrm{ball}(u)$ is inserted.

@@

We will make use of Assumption \ref{assumption:independence} in the analysis. 
\begin{assumption}
\label{assumption:independence}
All events represented by the nodes and edges of a witness tree are stochastically independent. 
\end{assumption}
In other words, we assume that no ball is referenced by more than one node of a witness tree. Obviously, a correct analysis will have to consider balls assigned to more than one node, but this simplifying assumption is useful to highlight the differences between the symmetric and asymmetric allocation schemes. Section \ref{sec:analysis:nondistinctBalls} will extend the analysis to witness trees with reduntant balls.

@@

An edge\footnote{where $v$ is the $i$'th child of $u$} $(u, v)$ is activated with probability at most $d/n$, which is the probability that the $i$'th location of the ball $b_u = \mathrm{ball}(u)$ points to a particular location of the ball $b_v = \mathrm{ball}(v)$ times the number of locations $d$. This implies a probability of at most
\begin{align}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align} 
or the event that all $k-1$ edges of a witness tree are activated. Note that  the assumption that all balls are distinct (Assumption \ref{assumption:independence}) implies that all edge events are independent.

>>>>

\usepackage{color}
\newcommand\todo[1]{\textcolor{red}{(TODO: #1)}}
\newcommand\load{L_{\mathrm{max}}}

@@

\title{The Power of Two Random Choices\\ 
\large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Tr\"aff and Francesco Versaci, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}}
\author{Martin Kalany, 0825673}

\maketitle
\begin{abstract}

@@

\section{Introduction}
\label{sec:intro}
Suppose that $m$ balls are thrown into $n$ bins, according to some arbitrary placement strategy. We are interested in the \emph{number of balls in the fullest bin}, also called the \emph{maximum load}. Equivalently, one can look at the \emph{additive gap} of the allocation, that is the difference of the number of balls in the fullest bin and the average amount of balls in the bins. These \emph{balls-into-bins games} or \emph{allocation problems} have been studied extensively in the probability literature (see e.g., \cite{JK77}). In the context of Computer Science, the balls-into-bins model is useful to study, e.g., strategies for distributing \emph{tasks} (balls) to \emph{processors} (bins) as evenly as possible without having to query all processors for the number of tasks currently assigned to them.

\subsection{Balls-into-bins model}
\label{sec:model}
Although several different variants of this simple balls-into-bins game exist, we will assume a very basic model to highlight the crucial ideas in both placement strategies as well as the analysis of them. In our model, we start with $n$ bins and put balls into them, one per time step $t$. We assume that at any time step $t$ at most $m = h n$ balls are in the bins, where $h$ is some constant greater or equal to $1$. At each time step $t$, we delete one randomly chosen ball from the bins. While most of the paper and especially the analysis focuses on the lightly loaded case where $m=n$, we will give results for the case $m > n$ as well. Throughout the paper, we will denote the maximum load by $\load$ and the additive gap by $\delta$. Obviously, 
\begin{align}
\delta = \load - \frac{m}{n}.
\end{align}

\subsection{Single-choice balls-into-bins}
\label{sec:single-choice}
A simple strategy for the balls-into-bins problem is to choose one bin uniformly and independently at random for each ball, which we will call the \emph{single-choice balls-into-bins} strategy. Using this strategy, the maximum load of the fullest bin will be
\begin{comment}
\footnote{A more accurate result exists: The maximum load is $\load = \Gamma^{-1}\left(n\right)\left(1+O\left(\frac{1}{\log \Gamma^{-1}\left(n\right)}\right)\right)$ \cite{G91}.}
\end{comment}
\begin{align}
\load = \frac{\ln n}{\ln \ln n} \cdot \left(1 + o\left(1\right)\right)
\end{align}
for the case $m=n$, with high probability\footnote{We say that an event $\mathcal E$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E \right] \geq 1 - n^{-\alpha}$ for an arbitrary constant $\alpha > 0$.}~\cite{RS98}. This implies\footnote{For a detailed proof, see Theorem \ref{theorem:asymptotic} in Appendix \ref{sec:asymptotic}.}
\begin{align}
\load\in \Theta\left(\frac{\ln n}{\ln \ln n}\right).
\end{align}
Note that since the average load is 1, $\delta = L_{max}$ in this case. For the heavily loaded case, w.h.p.~the maximum gap is at most~\cite{RS98}.
\begin{align}
\delta = \Theta\left(\sqrt{\frac{m \cdot \log n}{n}}\right).
\end{align}

\subsection{Multiple-choice balls-into-bins}
Consider a variation of the above \emph{single-choice balls-into-bins} strategy: For each ball, chose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}\footnote{If there are multiple bins with the same minimal load, chose one among them arbitrarily.}. This simple \emph{multiple-choice balls-into-bins} scheme, which we will call the \emph{Greedy} algorithm\footnote{A detailed definition will follow in Section \ref{sec:greedy}}, achieves a gap
\begin{align}
\delta = \frac{\log \log n}{\log d} + O(1)
\end{align}
 w.h.p.~for all $m$~\cite{ABKU99}~\cite{BCSV06}.

This apparently minor change introduces a small amount of choice which, compared to the single-choice scheme, leads to a significant reduction in the maximum load. 
\begin{comment}
Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. 
\end{comment}
Extensive research following the influential paper by Azar et al.~\cite{ABKU99}, who were the first to explicitly prove this phenomenon\footnote{For a brief history of the proof, which is quite interesting by itself, refer to Appendix \ref{sec:historyOfGreedy}.}, suggests that this \emph{two-choice paradigm} is not just an artifact of the simple balls-into-bins model, but a general principle applicable to a variety of problems.

Consider a slightly more involved strategy, which we will call the \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size\footnote{For the sake of a clear presentation, we will assume that $d$ divides $n$ evenly in the remainder of this paper.} $\lfloor \frac{n}{d}\rfloor$ or $\left\lceil\frac{n}{d}\right\rceil$. We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with the same minimal load, we put the ball into the bin with smallest index $i$. Somewhat surprisingly, V\"ocking \cite{VOC03} showed that such a strategy employing \emph{asymmetric tie-breaking} in combination with \emph{non-uniform} bin selection reduces the bound for the gap to\footnote{The symbol $\Phi_d$ denotes a generalization of the golden ratio, based on generalized Fibonacci numbers. Refer to Appendix \ref{sec:fibonacci} for the definition.}
\begin{align}
\delta = \frac{\ln\ln n}{d \cdot \ln \Phi_d} + O(1)
\end{align}
w.h.p. Not only is this a significant reduction of the asymptotic bound, V\"ocking was able to show that the Always-go-left strategy is optimal up to additive constants \cite{VOC03}.

This paper is organized as follows: We give a basic classification of balls-into-bins games in Section \ref{sec:classification}, to provide some overview of the different directions in research as well as possibilities of abstraction from applications, some of which are listed in Section \ref{sec:applications}. In the remaining sections of this paper we will consider the model given in Section \ref{sec:model}, which can be classified as a \emph{sequential, finite and dynamic process with random deletions}. Section \ref{sec:multiple-choice} starts with a brief definition of three different classes of random placement strategies and then continues with the presentation of the two placement algorithms outlines in the introduction: The  \emph{Greedy} algorithm (Section \ref{sec:greedy}) and the \emph{Always-go-left} algorithm (Section \ref{sec:AlwaysGoLeft}). Section \ref{sec:analysis} is the core of this paper and will give a detailed analysis of both strategies. We will use a proof technique called \emph{witness trees}, which allows for a clean analysis of both problems. Note that the technique is applicable to a wide variety of balls-into-bins scenarios, making it the most useful of several proof techniques applied to balls-into-bins problems. Other proof techniques such as \emph{layered induction} and a \emph{differential equations approach} have been used. They are however not part of this paper. See, e.g., \cite{MRS01} for an overview of different proof techniques for the balls-into-bins problem.

\begin{comment}
In Section \ref{sec:strategies}, we will provide asymptotic bounds for the ad dative gap of this strategy, which we will then use to devise more complex strategies that significantly reduce these bounds. 
\end{comment}

@@

\section{Classification}
\label{sec:classification}
Several different variations of this problem exist, which may be classified by the following properties:
\begin{compactitem}
\item In a \emph{sequential process}, the balls arrive one after another. A  ball is dealt with only after the previous ball has been placed. Contrary, in a \emph{parallel process}, balls arrive in sets and the balls in one set are dealt with simultaneously \cite{ABS98}. 
\item In a \emph{static problem}, the balls are placed into the bins and no balls are ever removed. In a \emph{dynamic problem}, balls can also be removed from the system. In a simple \emph{random deletions} model, for each newly inserted ball a ball chosen uniformly and independently at random from among the balls currently in the bins is removed from the system \cite{ABKU99} \cite{MRS01}. More elaborate schemes exists: An \emph{oblivious adversary} specifies the sequence of insertions and deletions of balls in advance, without any knowledge of the placement of the balls \cite{CFM+98}. Another approach was studied by Adler et al.~\cite{ABS98} in the context of the parallel balls-into-bins problem: In each bin, the balls are stored in FIFO-order and at every time step the first ball in each of the bins is deleted. 
\item A \emph{finite process} is analyzed for a given time interval that is known beforehand, whereas for an \emph{infinite process}, a placement and a deletion process is defined over an infinite time span \cite{ABS98}.
\item In the \emph{lightly loaded case}, $m$ is in the order of $n$, whereas in the \emph{heavily loaded case}, $m$ is arbitrarily large. For clarity, we will assume that $m = n$ in the lightly loaded case and that $m > n$ in the heavily loaded case. 
\end{compactitem}

@@

To maximize the throughput of the servers, a uniform distribution of tasks to servers is desirable. This can easily be achieved by a central dispatcher. However, such a central dispatcher poses a bottleneck within a distributed system and querying $m$ servers for their current load involves significant overhead, e.g., sending a message to each server and waiting for the reply and more efficient solutions are necessary. As with hashing, the balls-into-bins model can be applied easily: we need to assign tasks (the balls) to servers (the bins). Assuming tasks are assigned to servers uniformly and independently at random, the two-choice paradigm is directly applicable: If each client samples two random servers for their load and sends issues its request to the least loaded server, the overhead is comparatively small and the difference in load on the $n$ servers is bounded by $\Theta\left(\log \log n\right)$ w.h.p. \cite{KLM92} \cite{RS98} \cite{MRS01}.

\subsection{Emulation of shared memory machines on distributed memory machines}
\label{sec:DMM}
The \emph{Parallel Random Access Machine} model\footnote{as defined in the lecture or in~\cite{P03}}, or \emph{PRAM} for short, is a high level abstraction of a shared memory machine, in which memory access by any processor to any memory location is assumed to require constant time. It is especially useful for the study of parallel algorithms. However, a PRAM is much harder to realize in hardware than a \emph{Distributed Memory Machine} (DMM). Thus, ways to emulate a PRAM on a DMM have been studied extensively. Such an emulation distributes the processors and memory cells of the PRAM to the processors and memory modules (typically one per processor) of the DMM with the goal to reduce the slowdown of the emulation, which is the time the DMM requires to simulate one step of the PRAM. Each memory module can handle only one memory access at a time step. More accesses to the same module will result in memory contention. The balls-into-bins idea is useful for managing communication between the processors and the memory modules such that memory contention is minimal. For details, see Karp et al.~\cite{KLM92} and the papers referenced there.

\subsection{Low congestion circuit routing}
\label{sec:circuitrouting}
In multi-stage interconnection networks, virtual circuit-switching is used to route communication. For simplicity, we consider the \emph{permutation routing problem}, where one request originates at each of the $n$ input nodes and one request is destined for each of the $n$ output nodes and assume a \emph{butterfly network} as the underlying interconnection network (see \cite{CLR09} for a detailed definition). Then, congestion is the maximum number of paths that are routed through a network link. The goal of a \emph{circuit routing algorithm} is to to allocate paths for all $n$ communication requests such that congestion is minimal.

@@

@@

\section{Classes of random placement strategies}
\label{sec:classesOfPlacement}
Let $B$ be the set of bins $\{1,2...n\}$, $\Omega = B^{d}$ be the sample space and $\mathcal{F} = 2^{\Omega}$ be the set of measurable events. We are only interested in events that represent a set of bins of size $d$. Thus, let $\mathcal{F}_d = \{\mathcal{E} \in \mathcal{F}: \left\vert \mathcal{E} \right\vert = d\}$ and set $\Pr\left(\mathcal{E}  \right) = 0$  $\forall \mathcal{E} \in \mathcal{F} \setminus \mathcal{F}_d $. The three classes of algorithms are distinguished by the probability function $\Pr\left(\mathcal{E}\right)$ for events $\mathcal{E} \in \mathcal{F}_d$:
\begin{compactitem}
\item  \emph{Uniform and independent.} Choose each of the $d$ bins for a ball uniformly and independently at random from the set of bins $B$. This implies 
\begin{align}
\forall \mathcal{E} \in \mathcal{F}_d: \Pr\left(\mathcal{E}\right) = \frac{1}{n ^{d}}
\end{align}
\item Class 2: \emph{Non-uniform and independent.} Choose the i-th bin from $B$ independently at random according to some probability function
\begin{align}
\Pr: B \rightarrow \left[0,1\right]
\end{align}

@@

\begin{theorem}
\label{theorem:greedy}
Consider any sequence of insertions and deletions s.t.~at most $hn$ balls are in the bins at any time. If the balls are placed into bins by the Greedy algorithm, the maximum load is at most 
\begin{align}
\frac{m}{n} + \frac{\log \log n}{\log d}+ O(h)
\end{align}
w.h.p.~at any time $t$ \cite{ABKU99} \cite{BCSV06}.
\end{theorem}

@@

\subsection{The Always-go-left scheme}
\label{sec:AlwaysGoLeft}
The \emph{Greedy} strategy presented in Section \ref{sec:multiple-choice} assumes that $d$ bins are chosen uniformly and independently at random. V\"ocking \cite{VOC03} was the first to study the impact of non-uniform and dependent selections of the $d$ bins. Surprisingly, a multiple-choice balls-into-bins strategy with a \emph{non-uniform} and independent selection of the $d$ bins, such as the \emph{Always-go-left} algorithm introduced in \cite{VOC03}, achieves better \cite{BCSV06} load balancing than the \emph{Greedy} scheme (Section \ref{sec:greedy}). Furthermore, V\"ocking proved that the \emph{Always-go-left} algorithm is optimal up to additive constants.

@@

\begin{theorem}
\label{theorem:agln}
If $n$ balls are placed into $n$ bins using the Always-go-left scheme, the number of balls in the fullest bin will be 
\begin{align}
\frac{\ln\ln n}{d \cdot \ln \Phi_d} + O(1)
\end{align}
w.h.p.
\end{theorem}

@@

We say that a \emph{bad event} occurs when the maximum load exceeds some given threshold value. When such a bad event occurs we construct an \emph{activated witness tree}. In other words, the activation of a witness tree is implied by a bad event. The probability for the existence of such an activated witness tree is an upper-bound for the probability that the associated bad event occurs.

Throughout this section, \emph{location} will denote the a specific place in one of the $n$ bins. A location of a ball $b$ refers not only to a bin, but also to the number of balls already in this bin at the time of insertion of ball $b$.

We will prove the bounds for $m=n$ for both the Greedy and the Always-go-left scheme in Sections \ref{sec:analysis:greedy} and \ref{sec:analysis:alg}, respectively. However, we use a simplifying assumption \ref{assumption:independence} in those two sections to provide a more clear presentation of the proof technique. Section \ref{sec:analysis:nondistinctBalls} will extend these proofs to remove the simplifying assumption we make initially. The witness tree method can be used to show that the same bounds hold for the heavily loaded case where $m> n$ too, but we do not give such a proof in this paper. Instead, we refer the interested reader to V\"ocking \cite{VOC03}. Berenbrink et al.  ~\cite{BCSV06}, Talwar et al.~\cite{TW13} and Mitzenmacher \cite{MRS01} provide proofs the heavily loaded case using different approaches as well.

\subsection{Greedy}
\label{sec:analysis:greedy}
In this Section, we will give a detailed proof for Theorem \ref{theorem:greedy}. However, we will restrict ourselves to the case where $m = n$ and will make use of the symplifying assumption \ref{assumption:independence}.

\subsubsection{Symmetric witness trees}
\label{sec:analysis:definitionSymWT}
A \emph{symmetric witness tree} of order $L$ is a complete $d$-ary tree with $d^{L}$ leaves. Each node $v$ represents a ball $b_v = \mathrm{ball}(v)$ . The same ball may be represented by several nodes. The nodes and edges of the witness tree represent events that may occur or not. For simplicity, we assume that all events represented by a witness tree are stochastically independent.

We define edge and leaf events:
\begin{compactitem}
\item \emph{Edge event:} Consider some edge $e = (u,v)$ of the witness tree, s.t.~$v$ is the i-th child of $u$. Then $e$ represents the event that the i-th location of the ball $b_u = \mathrm{ball}(u)$ points to the same bin as one of the locations of the ball $b_v = \mathrm{ball}(v)$. 
\todo{illustrate}
\item \emph{Leaf event:} A leaf node $v$ of the witness tree represents the event that each of the $d$ locations of the ball $b_v = \mathrm{ball}(v)$ points to a bin that contains at least three additional balls (i.e., balls that are not represented by a node of the tree) at the time of the insertion of the ball $b_v$.
\todo{illustrate}
\end{compactitem}

The following constraints have to be fulfilled by an assignment. The ball $b_r$ that is represented by the root node $r$ has to exist at time step\footnote{As defined in Theorem \ref{theorem:greedy}} $t$ and each ball $b_v$ represented by a node $v \neq r$ and parent node $u$ has to exist when the ball $b_u = \mathrm{ball}(u)$ is inserted.

@@

Initially, we will make use of the following assumption in the analysis. 
\begin{assumption}
\label{assumption:independence}
All events represented by the nodes and edges of a witness tree are stochastically independent. 
\end{assumption}
In other words, we assume that no ball is referenced by more than one node of a witness tree. Obviously, a correct analysis will have to consider balls assigned to more than one node, but this simplifying assumption is useful to highlight the differences between the symmetric and asymmetric allocation schemes. Section \ref{sec:analysis:nondistinctBalls} will extend the analysis to witness trees with redundant balls.

@@

An edge\footnote{where $v$ is the i-th child of $u$} $(u, v)$ is activated with probability at most $d/n$, which is the probability that the i-th location of the ball $b_u = \mathrm{ball}(u)$ points to a particular location of the ball $b_v = \mathrm{ball}(v)$ times the number of locations $d$. This implies a probability of at most
\begin{align}
p_e = \left(\frac{d}{n}\right)^{k-1}
\end{align} 
or the event that all $k-1$ edges of a witness tree are activated. Note that  the assumption that all balls are distinct (Assumption \ref{assumption:independence}) implies that all edge events are independent.
